---
title: "Homework 3"
author: "Natalie Parker"
date: "05-28-26"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

# Set up

```{r}
#| label: libraries-and-data

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects)

# storing kelp data as object "kelp"
kelp <- read.csv(here("data", "temp-kelp.csv"))
# creating new object that fits a linear model to the data
kelp_model <- lm(
  kelp_elong ~ temp_c, 
  data = kelp)
# storing personal data as object "personal_data"
personal_data <- read_xlsx(here("data", "ENVS 193DS- Personal Data.xlsx"))
```

# Problem 1

## 1a.

The appropriate tests would be the Pearson's correlation and linear regression to determine the strength of the relationship between giant kelp frond elongation rate and temperature. Pearson's correlation measures the strength and the direction between the two variables. Linear regression aims to describe a linear relationship between the variables while being able to describe strength as well. 

## 1b. 
```{r}
#| label: visualization-of-temp-and-kelp
#| warning: false
# creating new object called kelp_preds
kelp_preds <- ggpredict(
  kelp_model, 
  terms = "temp_c")
# base layer: ggplot with kelp data
ggplot(data = kelp, 
       aes (x = temp_c, # x-axis is sea temp
            y = kelp_elong)) + # y-axis is kelp elongation rate
  # first layer: geom_point to show underlying data
  geom_point(fill = "#1f78b4", # changing the color and size
             size = 4, 
             stroke = 1,
             shape = 21) + 
 # second layer: ribbon representing confidence interval
 # using predictions data frame
 geom_ribbon(data = kelp_preds,
             aes(x = x,
                 y = predicted, 
                 ymin = conf.low,
                 ymax = conf.high),
             alpha = 0.1) +
  # third layer: line representing model predictions
  geom_line(data = kelp_preds,
            aes(x = x,
                y = predicted),
            linewidth = 1)+
   labs ( x= "Average Ocean Temperature (°C)",
         y = bquote(Frond~Elongation~Rate~(cm~day^-1)),
                    title = "Giant Kelp Elongation Rate vs. Ocean Temperature") +
  # changing the theme
  theme_minimal() +
  # customizing the look of the titles!
  theme(plot.title = element_text(face = "bold",
                                  size = 14, 
                                  hjust = 0.5),
        axis.title = element_text(face = "plain",
                                  size = 11))
```


## 1c. 

### Checking assumptions
```{r}
#| label: diagnostic-plots-to-check-assumptions
# base R residuals
par(mfrow = c(2,2))
plot(kelp_model)

```
When thinking about the relationship between ocean temperature and kelp elongation, it would make sense for there to be a linear relationship. We can see this in the residuals vs. fitted plot and the plotted data. 
We can assume independent errors due to the independent observations and data collection methods. 
The residual vs. fitted and scale location plots both show us the homoscedasticity of errors. The plots show that the data is homoscedastic because of the decently straight trend line. 
Looking at the QQ plot seen in residual plots, the data seems to be normally distributed. It roughly follows the straight line with variations that are normal to see in real data. 

I think I did ok checking the assumptions. It's hard because making the decision is subjective, but generally most math isn't, and making the "wrong" choice can have big impacts. The QQ plot was hardest to assess due to the slight curve seen in it.  

### Running test
```{r}
#| label: showing-results
# showing the output of linear regression
summary(kelp_model)
```

## 1d. 
To evaluate the strength of the relationship between temperature and giant kelp frond elongation rate, I used an F test from a linear regression model. I used the linear regression model because the data met the assumptions for the model and the goal was to test the strength of the relationship of the variables. A linear model is great for testing strength and relationship. I used the f-test because it's traditional to get the output for an F-test when creating a linear model. 

The elongation rate of giant kelp decreased as the average ocean temperature increased (linear regression, F(30,1) = 26.93, R^2^ = 0.473, p< 0.001, $\alpha$ = 0.05)

Based on Model predictions, for each unit 1 increase in ocean temperature, we expect an decrease of 0.43 $\pm$ 0.08 (SE) cm day^-1^.



