---
title: "Homework 3"
author: "Natalie Parker"
date: "05-28-26"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

# Set up

```{r}
#| label: libraries-and-data

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects)

# storing kelp data as object "kelp"
kelp <- read.csv(here("data", "temp-kelp.csv"))
# creating new object that fits a linear model to the data
kelp_model <- lm(
  kelp_elong ~ temp_c, 
  data = kelp)
# storing and cleaning personal data as object "personal_clean"
personal_clean <- read_csv(
  here("data", "ENVS 193DS- Personal Data - Sheet1.csv"), 
  # skipping the first 2 messed up rows
  skip = 2) |> 
  # dropping the first empty column
  select(-1) |> 
  # cleaning the names
  clean_names() |> 
 # converting sleep to a graphable value
  mutate(
    sleep_hours = 
      hour(hms(amount_of_sleep)) + (minute(hms(amount_of_sleep)) / 60),
    # converting the columns of "date", "caffeine", and "satisfaction" to numbers and dates
    date = as.Date(date, format = "%m/%d/%Y"),
    caffeine = as.numeric(caffeine),
    satisfaction = as.numeric(satisfaction))
```

# Problem 1

## 1a.

The appropriate tests would be the Pearson's correlation and linear regression to determine the strength of the relationship between giant kelp frond elongation rate and temperature. Pearson's correlation measures the strength and the direction between the two variables. Linear regression aims to describe a linear relationship between the variables while being able to describe strength as well. 

## 1b. 
```{r}
#| label: visualization-of-temp-and-kelp
#| warning: false
# creating new object called kelp_preds
kelp_preds <- ggpredict(
  kelp_model, 
  terms = "temp_c")
# base layer: ggplot with kelp data
ggplot(data = kelp, 
       aes (x = temp_c, # x-axis is sea temp
            y = kelp_elong)) + # y-axis is kelp elongation rate
  # first layer: geom_point to show underlying data
  geom_point(fill = "#1f78b4", # changing the color and size
             size = 4, 
             stroke = 1,
             shape = 21) + 
 # second layer: ribbon representing confidence interval
 # using predictions data frame
 geom_ribbon(data = kelp_preds,
             aes(x = x,
                 y = predicted, 
                 ymin = conf.low,
                 ymax = conf.high),
             alpha = 0.1) +
  # third layer: line representing model predictions
  geom_line(data = kelp_preds,
            aes(x = x,
                y = predicted),
            linewidth = 1)+
   labs ( x= "Average Ocean Temperature (°C)",
         y = bquote(Frond~Elongation~Rate~(cm~day^-1)),
                    title = "Giant Kelp Elongation Rate vs. Ocean Temperature") +
  # changing the theme
  theme_minimal() +
  # customizing the look of the titles!
  theme(plot.title = element_text(face = "bold",
                                  size = 14, 
                                  hjust = 0.5),
        axis.title = element_text(face = "plain",
                                  size = 11))
```


## 1c. 

### Checking assumptions
```{r}
#| label: diagnostic-plots-to-check-assumptions
# base R residuals
par(mfrow = c(2,2))
plot(kelp_model)

```
When thinking about the relationship between ocean temperature and kelp elongation, it would make sense for there to be a linear relationship. We can see this in the residuals vs. fitted plot and the plotted data. 
We can assume independent errors due to the independent observations and data collection methods. 
The residual vs. fitted and scale location plots both show us the homoscedasticity of errors. The plots show that the data is homoscedastic because of the decently straight trend line. 
Looking at the QQ plot seen in residual plots, the data seems to be normally distributed. It roughly follows the straight line with variations that are normal to see in real data. 

I think I did ok checking the assumptions. It's hard because making the decision is subjective, but generally most math isn't, and making the "wrong" choice can have big impacts. The QQ plot was hardest to assess due to the slight curve seen in it.  

### Running test
```{r}
#| label: showing-results
# showing the output of linear regression
summary(kelp_model)
```

## 1d. 
To evaluate the strength of the relationship between temperature and giant kelp frond elongation rate, I used an F test from a linear regression model. I used the linear regression model because the data met the assumptions for the model and the goal was to test the strength of the relationship of the variables. A linear model is great for testing strength and relationship. I used the f-test because it's traditional to get the output for an F-test when creating a linear model. 

We reject the null hypothesis, so the elongation rate of giant kelp decreased as the average ocean temperature increased (linear regression, F(30,1) = 26.93, R^2^ = 0.473, p< 0.001, $\alpha$ = 0.05)

Based on Model predictions, for each unit 1 increase in ocean temperature, we expect an decrease of 0.43 $\pm$ 0.08 (SE) cm day^-1^.

## 1e. 

This test used the data collected from kelp in varying ocean temperatures to understand how increasing ocean temperature can affect giant kelp growth. From performing a F test for significance and a linear regression model for strength of the relationship, we were able to understand that when average ocean temp increases, we see a decrease in giant kelp elongation. This means that for giant kelp to be at their healthiest, ocean temperatures cannot continue to increase. 

## 1f. 
```{r}
#| label: pearsons-correlation
# creating new object for pearson's test from kelp data and running pearson's test
kelp_corr <- cor.test(kelp$temp_c, 
                      kelp$kelp_elong, 
                      method = "pearson")
# displaying test output
kelp_corr
```

Both tests would have lead to the exact same decision of rejecting the null hypothesis because they have identical t-values of -5.189, p-values of 1.366e-05, and the same significance level of 0.05. They also come to the same conclusion for correlation and strength because of they have to same regression coefficient (-0.688), which is equal to the R squared value of 0.473 when squared. They both show a strong negative correlation between increase in temp and kelp elongation. 

# Problem 2

## 2a. 
```{r}
#| label: original-length-vs.-satisfaction
# base layer: ggplot
ggplot(data = personal_clean,
       # x-axis is length of wo, y-axis is satisfaction
       aes(x = length_of_wo,
           y = satisfaction)) +
  # first layer: point, points with a custom color
  geom_point(color = "cyan4", size = 4, alpha = 0.8) +
  # clean theme
  theme_bw() +
  # removing the excess grid marks
  theme(panel.grid.minor = element_blank())+
  # labels and title
  labs(
    x = "Length of Workout (hours)",
    y = "Satisfaction Score (1-5)",
    title = "Length of Workout Doesn't Influence Satisfaction")
```

```{r}
#| label: new-satisfaction-vs.-sleep
workout_model <- lm(satisfaction ~ sleep_hours, data = personal_clean)
personal_preds <- ggpredict(workout_model, 
                            terms = "sleep_hours")
#base layer: ggplot 
ggplot() +
  # second layer: geom_point to show underlying data with personal_clean data
  geom_point( data = personal_clean, 
       aes(x = sleep_hours, 
       y = satisfaction), 
       fill = "#ac88ff",
       size = 4,
       stroke = 1,
       shape = 21) +
  # third layer: geom_ribbon to show CI
 # using predictions data frame
 geom_ribbon(data = personal_preds,
             aes(x = x,
                 y = predicted, 
                 ymin = conf.low,
                 ymax = conf.high),
             fill = "#00bdd0",
             alpha = 0.1) +
  # third layer: line representing model predictions
  geom_line(data = personal_preds,
            aes(x = x,
                y = predicted),
            color = "#00bdd0",
            linewidth = 1) +
 
  # Clean labels with explicit units
  labs(
    x = "Amount of Sleep (Hours)",
    y = "Workout Satisfaction Score (1-5)",
    title = "Workout Satisfaction Predicted by Sleep Duration") +
  # Apply clean base theme and strip all grid lines
  theme_classic() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    axis.title = element_text(face = "plain", size = 11))


```

